Quasi-Monte Carlo Methods: What, Why, and How?

### 2025-01-27 HickernellMCQMC2024_QMCTutorial.tex
aleksei updates

	Choosing the nodes to lie on a grid corresponds to a tensor product midpoint cubature rule, which would normally be expected to have an error of $\mathcal{O}(n^{-2/d})$  for general $d$.  The error decay of $\mathcal{O}(n^{-1/5})$ rather than $\mathcal{O}(n^{-2/6})$ for this example may be due to the lack of smoothness of $f$ near the boundaries of the unit cube.
\begin{figure}
	\centering
	\includegraphics[width = 0.3\textwidth]{\figpath/64gridpts_d2.pdf}\qquad
	\includegraphics[width =  0.3\textwidth]{\figpath/64gridpts_d6.pdf}
	\caption{\FJHNote{Is the font size large enough?}Although a two-dimensional grid covers the unit square rather well, a $d$-dimensional grid for modest $d$ does \emph{not} cover the unit cube well.  For example, one can only see four distinct nodes in a two dimensional projection of a six-dimensional grid with $64$ nodes. \label{fig:grid}}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width =\textwidth]{\figpath/64iidpts_d6.pdf}
	\caption{IID nodes cover the unit cube better than grid nodes, although one does observe clusters and gaps.  In any  projection there is  a similar looking distribution of all $64$ nodes \label{fig:iid}}
\end{figure}

@@

The Halton construction is an extensible LD sequence. However, should one know $n \in \mathbb{N}$ in advance, a finite $d+1$-dimensional LD node set called the Hammersley node set can be defined similarly as
\begin{equation}\label{eq:hammersley}
    \bsx_i^{\text{Ham}} = \left( \frac{i}{n}, \phi_{b_1}(i), \ldots, \phi_{b_d}(i) \right) \quad i = 0, 1, \ldots, .
\end{equation}

@@

Most quasi-Monte Carlo constructions are studied based on their asymptotic behavior as \(n \to \infty\), with numerous results on asymptotic rates \cite{NIED1978QMCBook,NovWoz10a}. However, few methods address finding optimal node configurations for specific \(n\) and \(d\) which can be critical in practical applications where each function evaluation is quite costly. Of course, one possible strategy is to truncate asymptotically optimal sequences to the desired \(n\), however, as discussed above many of the standard LD constructions have a preferred number of nodes with truncations leading to imbalances in equidistribution. We can often do better optimizing directly for the target \(n\) and \(d\).

@@

\begin{figure}
	\centering
	\includegraphics[width = \textwidth]{\figpath/n_8_n_scr_3_scrambled_sobol_sequence_d2.pdf}
	\caption{\FJHNote{Something went awry here.}Three linear scrambles of a digital $(0,3,2)$-net with nodes denoted by three different colors.  There is one node in each smaller box of the form \eqref{eq:element_box} for $\lVert \bsk \rVert = 3$, for each digitally shifted net. \label{fig:scrambled_net}}
\end{figure}

\subsection{Other Randomizations for Halton}

@@

Instead of explicitly searching for good generating vectors or matrices, one can instead sample a few random generating vectors or generating matrices and then prune out bad choices \AGSNote{citation?}.  Random generating vectors for lattices take $\bsh \overset{\text{IID}}{\sim} \calU\{1,\dots,n-1\}^d$. Random generating matrices for base $2$ digital nets---as described in \eqref{eq:digitalB}---set $x^{\text{dig}}_{ij\ell} \overset{\text{IID}}{\sim} \calU\{0,1\}$ for $i<\ell$ while enforcing $x_{iji}=1$ and $x_{ij\ell}=0$ for $i>\ell$.

@@

A straightforward qMC stopping criterion may also be derived from the CLT using several randomizations of a LD node set. Let $\bsx_{ir}$ be the $i^\mathrm{th}$ node in the $r^\mathrm{th}$ randomization of a LD node set with $0 \leq i < n$ and $1 \leq r \leq R$. For example, we may use $R$ independent shifts of a lattice node set or independent digital shifts of a digital net. Crucially, the randomized node sets presented in Section \ref{sec:random} are also LD. \FJHNote{Tried to make the notation more consistent.}Let $\hat{\mu}^{(r)} = n^{-1} \sum_{i=0}^{n-1} f(\bsx_{ir})$ be sample means for each randomization and set $\hat{\mu}_{nR} = R^{-1} \sum_{r=1}^R \hat{\mu}^{(r)}$ to be the aggregate approximation. In the spirit of CLT, a standard approximate $1-\alpha$ confidence interval $(\mu^-,\mu^+)$ has $\mu^\pm = \hat{\mu}_{nR} \pm t_{R-1}^{1-\alpha/2} S/\sqrt{R}$ where $S^2 = (R-1)^{-1} \sum_{r=1}^R (\hat{\mu}^{(r)} - \hat{\mu}_{nR})^2$ and $t_{R-1}^{1-\alpha/2}$ is the $1-\alpha/2$ quantile of the Student $t$ distribution with $R-1$ degrees of freedom. Here we have used the Student $t$ distribution since $R$ is typically small, for example $R=15$. To use this confidence interval for a stopping criterion, one may iteratively double the sample size $n$ until the interval width is sufficiently low. In \cite{l2023confidence}, comprehensive numerical experiments show that this straightforward qMC stopping criterion achieve better coverage than more involved bootstrap $t$ methods across a variety of test functions.

@@

For this Bayesian approach, one must find the eigenvalues and eigenvectors of the Gram matrix $\mathsf{K} : = \bigl(K(\bsx_i,\bsx_j) \bigr)_{i,j=0}^{n-1}$.  In general this requires $\mathcal{O}(n^3)$ operations.  However, pairing shifted lattices with shift-invariant kernels as defined in \eqref{eq:shinvK} yields circulant Gram matrices for which computations can be done quickly using the fast Fourier transform \cite{RatHic19a}. Similarly, pairing digitally shifted digital nets with digitally-shift-invariant kernels yields recursive-block Toeplitz Gram matrices \AGSNote{Different name?} for which computations can be done quickly using the fast Walsh-Hadamard transform \cite{JagHic22a}.

@@

The above formulation is equivalent to importance sampling \AGSNote{reference} by a distribution with density $\nu$. Optimally, one would choose $\nu \propto g\lambda$ so that $f$ is constant, but this is typically infeasible in practice due to our limited knowledge of $g$. Choosing $\bsPsi$ so that $\bsT = \bsPsi(\bsX)$ gives the simplification $f(\bsX) = g(\bsPsi(\bsX))$. An example of this was given in Section \ref{sec:practice} where $\bsT \sim \calN(\boldsymbol{0},\mathsf{I}/2)$ and $\bsPsi(\bsX) = \Phi^{-1}(\bsX)/\sqrt{2}$ where $\Phi^{-1}$ is the inverse CDF of the standard Gaussian distribution applied elementwise.

@@

Let $\{Y_l := f_l(\bsX^{(l)})\}_{l=1}^L$ be a sequence of approximations to $Y = f(\bsX)$, where $\bsX^{(l)} = (X_1, \ldots, X_l) \sim \mathcal{U}[0,1)^{d_l}$.  Here we assume that $d_1 < \cdots < d_L$, and that the cost of evaluating $f_l(\bsx^{(l)})$ is $\mathcal{O}(d_l)$, where $\bsx^{(l)} = (x_1, \ldots, x_l)$.  Also, $d_L = d$, $Y_{d_L} = Y$, and $f_L = f$. Then the population mean may written as a telescoping sum:
\begin{align*}
    \mu & := \mathbb{E}(Y) = \mathbb{E}\bigl[f(\bsX)\bigr]
    = \mathbb{E}(Y_1) + \mathbb{E}(Y_2 - Y_1) + \cdots + \mathbb{E}(Y_{d_L} - Y_{d_{L-1}}) \\
    & = \int_{[0,1)^{d_1}} f_1(\bsx^{(1)}) \, \mathrm{d} \bsx^{(1)} +  \int_{[0,1)^{d_2}} \bigl[f_2(\bsx^{(2)}) - f_1(\bsx^{(1)})\bigr] \, \mathrm{d} \bsx^{(2)} + \cdots \\
    & \qquad \qquad + \int_{[0,1)^{d_L}} \bigl[f_L(\bsx^{(L)}) - f_{L-1}(\bsx^{(L-1)}) \bigr] \, \mathrm{d} \bsx^{(L)}.
\end{align*}
This series of integrals may be approximated by a series of sample means, namely
\begin{multline} \label{eq:MLmuhat}
    \hat{\mu}^{\text{ML}}_{\bsn,\bsd}  := \frac{1}{n_1}  \sum_{i=0}^{n_1-1} f_1(\bsx^{(1)}_i)  +  \frac{1}{n_2}  \sum_{i=0}^{n_2-1}  \bigl[f_2(\bsx^{(2)}_i) - f_1(\bsx^{(1)}_i)\bigr] + \cdots \\
     + \frac{1}{n_L}  \sum_{i=0}^{n_L-1}  \bigl[f_L(\bsx_i^{(L)}) - f_{L-1}(\bsx_i^{(L-1)}) \bigr].
\end{multline}
If $C_1$ is the computational cost of $f^{(1)}(\bsx^{(1)})$ and $C_l$ is the computational cost of \AGSNote{$f^{(l)}(\bsx^{(1)})$ of $f_L(\bsx_i^{(l)}) - f_{l-1}(\bsx_i^{(l-1)})$} for $l = 2, \ldots, L$, then the total computational cost of 
\eqref{eq:MLmuhat} is 
\begin{equation} \label{eq:MLcostdef}
    \text{cost}(\bsn,\bsd) = n_1 C_1 + \cdots + n_L C_L.
\end{equation}

@@

@@

Taking a broader view, Sorokin and Rao \cite{SorRao23a} illustrate that integration of qMC methods with Gaussian process models enhance the reliability of estimating failure probabilities in the machine learning domain. Furthermore, novel approaches for training deep neural networks utilizing LD nodes have been proposed in recent works \cite{longo21,mishra21}.
\FJHNote{Should include the water paper of Sorokin.  Also, these two sentences do not fit together.}

@@

%\section{Ongoing Research} \label{sec:ongoing}
\section{Conclusion} \label{sec:conclusion}
Quasi-Monte Carlo methods, based on low discrepancy sequences, can approximate the mean of a function, $f$, of $d$ variables to within an error tolerance of $\varepsilon$ at a cost of roughly $\mathcal{O}(\varepsilon^{-1})$ function evaluations, provided that the problem is formulated so that $f$ depends primarily on a modest number of variables.  This means that qMC outperforms alternative methods such as simple Monte Carlo and sampling on grids.  Using the multilevel qMC, the cost to reach the error tolerance is often essentially the same for $d=1$ and $d$ unbounded.  Software libraries are available to facilitate qMC computations.

\FJHNote{Our citations need to have Monte Carlo, Halton, etc. capitalized.}
\bibliographystyle{spmpsci}
\bibliography{FJH23,FJHown23,main}

>>>>

	Choosing the nodes to lie on a grid corresponds to a tensor product midpoint cubature rule, which would normally be expected to have an error of $\mathcal{O}(n^{-2/d})$  for general $d$.  The error decay of $\mathcal{O}(n^{-1/5})$ rather than $\mathcal{O}(n^{-2/6})$ for this example may be due to the lack of smoothness of $f$ near the boundaries of the unit cube.
\begin{figure}
	\centering
	\includegraphics[width = 0.3\textwidth]{\figpath/64gridpts_d2.pdf}\qquad
	\includegraphics[width =  0.3\textwidth]{\figpath/64gridpts_d6.pdf}
	\caption{Although a two-dimensional grid covers the unit square rather well, a $d$-dimensional grid for modest $d$ does \emph{not} cover the unit cube well.  For example, one can only see four distinct nodes in a two dimensional projection of a six-dimensional grid with $64$ nodes. \label{fig:grid}}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width =\textwidth]{\figpath/64iidpts_d6.pdf}
	\caption{IID nodes cover the unit cube better than grid nodes, although one does observe clusters and gaps.  In any  projection there is  a similar looking distribution of all $64$ nodes. \label{fig:iid}}
\end{figure}

@@

The Halton construction is an extensible LD sequence. However, should one know $n \in \mathbb{N}$ in advance, a finite $d+1$-dimensional LD node set called the Hammersley node set can be defined similarly as
\begin{equation}\label{eq:hammersley}
    \bsx_i^{\text{Ham}} = \left( \frac{i}{n}, \phi_{b_1}(i), \ldots, \phi_{b_d}(i) \right) \quad i = 0, \ldots, n-1.
\end{equation}

@@

Most quasi-Monte Carlo constructions are studied based on their asymptotic behavior as \(n \to \infty\), with numerous results on asymptotic rates \cite{Nie72,Nie92,NovWoz10a}. However, few methods address finding optimal node configurations for specific \(n\) and \(d\) which can be critical in practical applications where each function evaluation is quite costly. Of course, one possible strategy is to truncate asymptotically optimal sequences to the desired \(n\), however, as discussed above many of the standard LD constructions have a preferred number of nodes with truncations leading to imbalances in equidistribution. We can often do better optimizing directly for the target \(n\) and \(d\).

@@

\begin{figure}
	\centering
	\includegraphics[width = \textwidth]{\figpath/n_8_n_scr_3_scrambled_sobol_sequence_d2.pdf}
	\caption{Three linear scrambles of a digital $(0,3,2)$-net with nodes denoted by three different colors.  There is one node in each smaller box of the form \eqref{eq:element_box} for $\lVert \bsk \rVert = 3$, for each digitally shifted net. \label{fig:scrambled_net}}
\end{figure}

\subsection{Other Randomizations for Halton \FJHNote{Nathan will tackle this subsection and next}}

@@

Instead of explicitly searching for good generating vectors or matrices, one can instead sample a few random generating vectors or generating matrices and then prune out bad choices \cite{PanOwe23a}.  Random generating vectors for lattices take $\bsh \overset{\text{IID}}{\sim} \calU\{1,\dots,n-1\}^d$. Random generating matrices for base $2$ digital nets---as described in \eqref{eq:digitalB}---set $x^{\text{dig}}_{ij\ell} \overset{\text{IID}}{\sim} \calU\{0,1\}$ for $i<\ell$ while enforcing $x_{iji}=1$ and $x_{ij\ell}=0$ for $i>\ell$.

@@

A straightforward qMC stopping criterion may also be derived from the CLT using several randomizations of a LD node set. Let $\bsx_{ir}$ be the $i^\mathrm{th}$ node in the $r^\mathrm{th}$ randomization of a LD node set with $0 \leq i < n$ and $1 \leq r \leq R$. For example, we may use $R$ independent shifts of a lattice node set or independent digital shifts of a digital net. Crucially, the randomized node sets presented in Section \ref{sec:random} are also LD. Let $\hat{\mu}^{(r)} = n^{-1} \sum_{i=0}^{n-1} f(\bsx_{ir})$ be sample means for each randomization and set $\hat{\mu}_{nR} = R^{-1} \sum_{r=1}^R \hat{\mu}^{(r)}$ to be the aggregate approximation. In the spirit of CLT, a standard approximate $1-\alpha$ confidence interval $(\mu^-,\mu^+)$ has $\mu^\pm = \hat{\mu}_{nR} \pm t_{R-1}^{1-\alpha/2} S/\sqrt{R}$ where $S^2 = (R-1)^{-1} \sum_{r=1}^R (\hat{\mu}^{(r)} - \hat{\mu}_{nR})^2$ and $t_{R-1}^{1-\alpha/2}$ is the $1-\alpha/2$ quantile of the Student $t$ distribution with $R-1$ degrees of freedom. Here we have used the Student $t$ distribution since $R$ is typically small, for example $R=15$. To use this confidence interval for a stopping criterion, one may iteratively double the sample size $n$ until the interval width is sufficiently low. In \cite{l2023confidence}, comprehensive numerical experiments show that this straightforward qMC stopping criterion achieve better coverage than more involved bootstrap $t$ methods across a variety of test functions.

@@

For this Bayesian approach, one must find the eigenvalues and eigenvectors of the Gram matrix $\mathsf{K} : = \bigl(K(\bsx_i,\bsx_j) \bigr)_{i,j=0}^{n-1}$.  In general this requires $\mathcal{O}(n^3)$ operations.  However, pairing shifted lattices with shift-invariant kernels as defined in \eqref{eq:shinvK} yields circulant Gram matrices for which computations can be done quickly using the fast Fourier transform \cite{RatHic19a}. Similarly, pairing digitally shifted digital nets with digitally-shift-invariant kernels yields recursive-block Toeplitz Gram matrices for which computations can be done quickly using the fast Walsh-Hadamard transform \cite{JagHic22a}.

@@

The above formulation is equivalent to importance sampling by a distribution with density $\nu$. Optimally, one would choose $\nu \propto g\lambda$ so that $f$ is constant, but this is typically infeasible in practice due to our limited knowledge of $g$. Choosing $\bsPsi$ so that $\bsT = \bsPsi(\bsX)$ gives the simplification $f(\bsX) = g(\bsPsi(\bsX))$. An example of this was given in Section \ref{sec:practice} where $\bsT \sim \calN(\boldsymbol{0},\mathsf{I}/2)$ and $\bsPsi(\bsX) = \Phi^{-1}(\bsX)/\sqrt{2}$ where $\Phi^{-1}$ is the inverse CDF of the standard Gaussian distribution applied elementwise.

@@

Let $\{Y_l := f_l(\bsX^{(l)})\}_{l=1}^L$ be a sequence of approximations to $Y = f(\bsX)$, where $\bsX^{(l)} = (X_1, \ldots, X_l) \sim \mathcal{U}[0,1)^{d_l}$.  Here we assume that $d_1 < \cdots < d_L$, and that the cost of evaluating $f_l(\bsx^{(l)})$ is $\mathcal{O}(d_l)$, where $\bsx^{(l)} = (x_1, \ldots, x_l)$.  Also, $d_L = d$, $Y_{d_L} = Y$, and $f_L = f$. Then the population mean may written as a telescoping sum:
\begin{align*}
    \mu & := \mathbb{E}(Y) = \mathbb{E}\bigl[f(\bsX)\bigr]
    = \mathbb{E}(Y_1) + \mathbb{E}(Y_2 - Y_1) + \cdots + \mathbb{E}(Y_{d_L} - Y_{d_{L-1}}) \\
    & = \int_{[0,1)^{d_1}} f_1(\bsx^{(1)}) \, \mathrm{d} \bsx^{(1)} +  \int_{[0,1)^{d_2}} \bigl[f_2(\bsx^{(2)}) - f_1(\bsx^{(1)})\bigr] \, \mathrm{d} \bsx^{(2)} + \cdots \\
    & \qquad \qquad + \int_{[0,1)^{d_L}} \bigl[f_L(\bsx^{(L)}) - f_{L-1}(\bsx^{(L-1)}) \bigr] \, \mathrm{d} \bsx^{(L)}.
\end{align*}
This series of integrals may be approximated by a series of sample means, namely
\begin{multline} \label{eq:MLmuhat}
    \hat{\mu}^{\text{ML}}_{\bsn,\bsd}  := \frac{1}{n_1}  \sum_{i=0}^{n_1-1} f_1(\bsx^{(1)}_i)  +  \frac{1}{n_2}  \sum_{i=0}^{n_2-1}  \bigl[f_2(\bsx^{(2)}_i) - f_1(\bsx^{(1)}_i)\bigr] + \cdots \\
     + \frac{1}{n_L}  \sum_{i=0}^{n_L-1}  \bigl[f_L(\bsx_i^{(L)}) - f_{L-1}(\bsx_i^{(L-1)}) \bigr].
\end{multline}
If $C_1$ is the computational cost of $f^{(1)}(\bsx^{(1)})$ and $C_l$ is the computational cost of $f^{(l)}(\bsx^{(1)}) - f_{l-1}(\bsx_i^{(l-1)})$ for $l = 2, \ldots, L$, then the total computational cost of 
\eqref{eq:MLmuhat} is 
\begin{equation} \label{eq:MLcostdef}
    \text{cost}(\bsn,\bsd) = n_1 C_1 + \cdots + n_L C_L.
\end{equation}

@@

QMC methods have also found recent applications in scientific machine learning of PDE solutions. It has been shown that LD nodes are more efficient than IID nodes for training neural networks modeling PDE solution \cite{longo21,mishra21}. LD nodes have also been used to more efficiently fit Gaussian process regression models for operator learning the solution process of PDEs with uncertain coefficients  \cite{sorokin2024computationally}. There the computational speedup follows from pairing LD nodes with special kernels to induce nicely structured gram matrices, as used in the fast Bayesian cubature stopping criteria discussed in Section \ref{sec:stop}.

@@

@@

%\section{Ongoing Research} \label{sec:ongoing}
\section{Conclusion} \label{sec:conclusion}
Quasi-Monte Carlo methods, based on low discrepancy sequences, can approximate the mean of a function, $f$, of $d$ variables to within an error tolerance of $\varepsilon$ at a cost of roughly $\mathcal{O}(\varepsilon^{-1})$ function evaluations, provided that the problem is formulated so that $f$ depends primarily on a modest number of variables.  This means that qMC outperforms alternative methods such as simple Monte Carlo and sampling on grids.  QMC is particularly powerful when the effective dimension of the problem is small, even though the nominal dimension, $d$, may be large.  Software libraries are available to facilitate qMC computations.

\FJHNote{Our citations need to have Monte Carlo, Halton, etc. capitalized.}
\bibliographystyle{spmpsci}
\bibliography{FJH25,FJHown23,main}
